Multi-Agent Pathfinding with Matching using Increasing Cost Tree Search

### 2021-06-26 docs/paper.tex
Described matching generation strategy

			\EndProcedure
		\end{algorithmic}
		\caption{Increasing Cost Tree Search}
		\label{icts-algo}
	\end{algorithm}
	\subsection{Exhaustive ICTS}
	\label{exhaustive}
	For $K$ teams $t_1,\ldots,t_K$ of size $k_i$, there are in total $M = \prod_{i} k_i!$ matchings from team agents to team goals. Let $C_1$ through $C_M$ denote the optimal SIC costs of each MAPF instance corresponding to the MAPFM instance, with $C_m = \infty$ for infeasible $m$. In order to find the optimal solution, $m$ with lowest $C_m$ has to be found, which requires all $M$ to be solved or found to be infeasible, meaning that $\mathcal{O}(M)$ MAPF instances have to be solved. 
	\subsubsection{Bounded search}
	Once a solution is found for a matching $m$ with cost $C_m$, this can be used in the next search as an (exclusive) upper-bound $B$ for the SIC of nodes that have to be searched. Formally, let $B_0 = \infty$ represent the initial SIC bound of the ICT search. After solving each $m$ with solution cost $C_m$, let $B_{m+1} = \min(B_m,C_m)$. By successively updating this $B$ in this manner, as more matchings are solved, the effort needed to solve the remaining matchings can be expected to be reduced significantly. With $C_m$ denoting the optimal SIC for matching $m$, this effort can be expressed as	
	\[\Delta_{m,\text{bounded}} = \min(B_m,C_m) - C^*_m = B_{m+1} - C^*_m\]
	This is proportional to the effort due to the complexity of the ICTS search for matching $m$, $\mathcal{O}(k^{\Delta_{m,\text{bounded}}})$.
	If $B_m < C^*_m$, only the root node will be generated to determine this so $\Delta_{m,\text{bounded}} = 0$.
	Using this, the number of ICT nodes searched in total becomes $\mathcal{O}(\sum_m k^{\Delta_{m,\text{bounded}}})$.
	\subsubsection{Ordered enumeration}
	\label{ordered-enum}
	To reduce the total effort, the matchings $m$ can be enumerated and solved in increasing order of $C^*_m$. In this case, the effort per matching is monotonically non-increasing by Property \ref{monotonicity}. This induces a case in which this enumeration minimizes the number of nodes searched in total (Property \ref{cond-monoton-optim}). However, cases in which $\Delta_m$ of the first $m$'s searched is much higher could also be constructed, in which case this approach might cause more nodes to be processed in total than alternative methods. Therefore, this ordered enumeration remains a heuristic improvement that works well in practice, as shown in Section \ref{experiments}.

	\begin{property}[Monotonicity]
		\label{monotonicity}
		In bounded exhaustive search using ICTS, solving the matchings in increasing order of root SIC makes the effort $\Delta_{m,\text{bounded}}$ monotonically non-increasing as a function of $m$.
	\end{property}
	\begin{proof}
		Due to the order of matchings for successive matchings $m,m+t$ for some $t\geq 1$, $C^*_{m} \leq C^*_{m+t}$. By definition, $B_{m+2} = \min(B_{m+1},C_{m+1})$ meaning that $B_{m+2} \leq B_{m + 1}$ so transitively $B_{m+t+1} \leq B_{m+1}$. Therefore
		\[B_{m+t+1} - C^*_{m+t}\leq B_{m+1} - C^*_m\implies \Delta_{m+t,\text{bounded}} \leq \Delta_{m,\text{bounded}}\]
	\end{proof}
	\begin{property}[Conditional optimality]
		\label{cond-monoton-optim}
		Assuming that $\Delta_m$, the unbounded effort, is monotonically non-decreasing as a function of $m$ (with $m$ in increasing order of $C^*_m$), ordered enumeration minimizes the total number of nodes searched across matchings.
	\end{property}
	\subsubsection{Approximate ordered enumeration}
	In practice, it is difficult to enumerate all matchings in increasing order of root SIC.

	Finding a lowest SIC $m$ reduces to the assignment problem that can be solved using the Hungarian algorithm \cite{kuhn55} in polynomial time;

	however, to the knowledge of the author, no algorithm exists to more generally compute the $i$'th matching in increasing order of $C^*_m$ (assuming some fixed tie-breaking). This means that matchings have to be exhaustively enumerated and stored in memory to sort them, but this requires $\mathcal{O}(M)$ memory which was found to be more problematic in practice than the time complexity of the sorting, $\mathcal{O}(M\log(M))$.

@@

	In the implementation, ordered enumeration is approximated using a priority queue that maintains a subset of matchings of a fixed size $M_{pq}$\footnote{$M_{pq} = 10000$ was used in implementation but this could be much increased at the cost of memory}. Algorithm \ref{pq-algo} outlines the procedure.

@@

	CBM solving all problems with one team (Figure \ref{fig:r-25-1-p} and \ref{fig:r-75-1-p}), problems equivalent to anonymous MAPF instances, is at first surprising but follows directly from the definition of the CBM method. CBM solves single teams as anonymous MAPF instances in polynomial time, meaning that solving a single team reduces to a polynomial-time procedure, namely SSP. Following the results reported in \cite{ma2016}, it is expected that as the number of teams approaches $k$, CBM performance will deteriorate and thus become more similar to the results for exhaustive methods here reported.
	%	Still, some implementations might be more optimized than others so this might be a factor in performance characteristics and in particular in relative performance differences.
	\section{Conclusions and Future work}
	\label{conclusions}
	\label{future}
	The goal of this research was to seek alternative approaches to solving MAPFM based on ICTS. In this, the research has succeeded, but the scalability of the algorithms leaves something to be desired. Exhaustive methods, like those discussed in detail in Subsection \ref{exhaustive}, cannot be expected to scale well as teams grow larger due to their complexity, no matter what optimizations are put in place. Nonetheless, as was shown in Section \ref{experiments}, the more advanced exhaustive ICTS algorithms consistently outperform the ICTS-m based algorithms. Compared to the SIC-adaptation of the CBM method \cite{baauw2021}, it was seen that exhaustive methods generally perform relatively poorly if there are fewer teams and fewer obstacles, while being competitive once teams grow smaller and maps become more constrained. Delineating under which conditions exhaustive methods outperform CBM and vice versa is something for further research.

	%\section{Future work}
	A key idea for future research to look into is how to exploit the lower-bound on solution SIC that can be found using the Hungarian Algorithm \cite{kuhn55}. This lower-bound could be used in a branch-and-bound \cite{land1960} search of matchings. The most straightforward way would be to use an algorithm for MAPF at leaf nodes but algorithms for MAPFM could be used to solve (small) branches. Another possibility would be to use this bound to improve ICTS-m by only executing the $k$-agent low-level search for nodes that satisfy this bound. Lower-bounding in ICTS-m could be further improved by finding the optimal SIC for the anonymous MAPF instance corresponding to each team using a similar approach as \cite{baauw2021}. As long as there is a team that does not meet its bound, branching only on agents within this team is possible without losing optimality.

	A related observation is that the algorithms introduced in this work do not rely on the property of agents and goals having one color. Both exhaustive ICTS and ICTS-m could easily be adapted to support one or more colors (teams) for each agent and goal while still guaranteeing optimality, in contrast to CBM \cite{ma2016} and its SIC adaptation \cite{baauw2021} which specifically use this property to plan single teams in polynomial time. Future research could study the multi-color MAPFM variant and the application of methods described in this work to this problem.

>>>>

			\EndProcedure
		\end{algorithmic}
		\caption{Increasing Cost Tree Search}
		\label{icts-algo}
	\end{algorithm}
	\subsection{Exhaustive ICTS}
	\label{exhaustive}
	For $K$ teams $t_1,\ldots,t_K$ of size $k_i$, there are in total $M = \prod_{i} k_i!$ matchings from team agents to team goals. These matchings can be enumerated by indexing the $k_i!$ permutations of goals for each team $t_i$ as $1,\ldots,k_i!$. Using such indexing, each $e \in \{1,\ldots,k_1!\} \times \cdots \times \{1,\ldots,k_K!\}$ corresponds to a unique combination of permutations and can be mapped to a matching $m$ assuming some fixed ordering of agents within each team. Let $C_1$ through $C_M$ denote the optimal SIC costs of each MAPF instance corresponding to the MAPFM instance, with $C_m = \infty$ for infeasible matching $m$. In order to find the optimal solution, $m$ with lowest $C_m$ has to be found, which requires all $M$ matchings to be solved or found to be infeasible, meaning that $\mathcal{O}(M)$ MAPF instances have to be solved. 
	\subsubsection{Bounded search}
	Once a solution is found for a matching $m$ with cost $C_m$, this can be used in the next search as an (exclusive) upper-bound $B$ for the SIC of nodes that have to be searched. Formally, let $B_0 = \infty$ represent the initial SIC bound of the ICT search. After solving each $m$ with solution cost $C_m$, let $B_{m+1} = \min(B_m,C_m)$. By successively updating this $B$ in this manner, as more matchings are solved, the effort needed to solve the remaining matchings can be expected to be reduced significantly. With $C_m$ denoting the optimal SIC for matching $m$, this effort can be expressed as	
	\[\Delta_{m,\text{bounded}} = \min(B_m,C_m) - C^*_m = B_{m+1} - C^*_m\]
	This is proportional to the effort due to the complexity of the ICTS search for matching $m$, $\mathcal{O}(k^{\Delta_{m,\text{bounded}}})$.
	If $B_m < C^*_m$, only the root node will be generated to determine this so $\Delta_{m,\text{bounded}} = 0$.
	Using this, the number of ICT nodes searched in total becomes $\mathcal{O}(\sum_m k^{\Delta_{m,\text{bounded}}})$.
	\subsubsection{Ordered enumeration}
	\label{ordered-enum}
	To reduce the total effort, the matchings $m$ can instead be enumerated and solved in increasing order of $C^*_m$. In this case, the effort per matching is monotonically non-increasing by Property \ref{monotonicity}. This induces a case in which this enumeration minimizes the number of nodes searched in total (Property \ref{cond-monoton-optim}). However, cases in which $\Delta_m$ of the first $m$'s searched is much higher could also be constructed, in which case this approach might cause more nodes to be processed in total than alternative methods. Therefore, this ordered enumeration remains a heuristic improvement that works well in practice, as shown in Section \ref{experiments}.

	\begin{property}[Monotonicity]
		\label{monotonicity}
		In bounded exhaustive search using ICTS, solving the matchings in increasing order of root SIC makes the effort $\Delta_{m,\text{bounded}}$ monotonically non-increasing as a function of $m$.
	\end{property}
	\begin{proof}
		Due to the order of matchings for successive matchings $m,m+t$ for some $t\geq 1$, $C^*_{m} \leq C^*_{m+t}$. By definition, $B_{m+2} = \min(B_{m+1},C_{m+1})$ meaning that $B_{m+2} \leq B_{m + 1}$ so transitively $B_{m+t+1} \leq B_{m+1}$. Therefore
		\[B_{m+t+1} - C^*_{m+t}\leq B_{m+1} - C^*_m\implies \Delta_{m+t,\text{bounded}} \leq \Delta_{m,\text{bounded}}\]
	\end{proof}
	\begin{property}[Conditional optimality]
		\label{cond-monoton-optim}
		Assuming that $\Delta_m$, the unbounded effort, is monotonically non-decreasing as a function of $m$ (with $m$ in increasing order of $C^*_m$), ordered enumeration minimizes the total number of nodes searched across matchings.
	\end{property}
	\subsubsection{Approximate ordered enumeration}
	In implementation matchings were enumerated using a product of goal permutation indices as described above so they are not enumerated in increasing order of root SIC. As a result, to achieve this ordering, matchings have to be exhaustively enumerated and stored in memory in order to be sorted. This requires $\mathcal{O}(M)$ memory which was found to be more problematic in practice than the time complexity of the sorting, $\mathcal{O}(M\log(M))$.

	%Using the 
	%To generate matchings 
	%In the implementation of exhaustive matching, the algorithm used to generate matchings
	%The algorithm for generating 
	%In the implementation of ordered enumeration as described above, matchings are not generated in increasing order of root SIC. As a result, to achieve this ordering, matchings have to be exhaustively enumerated and stored in memory in order to be sorted. This requires $\mathcal{O}(M)$ memory which was found to be more problematic in practice than the time complexity of the sorting, $\mathcal{O}(M\log(M))$.

	%In practice, it is difficult to enumerate all matchings in increasing order of root SIC. Finding a lowest SIC $m$ reduces to the assignment problem that can be solved using the Hungarian algorithm \cite{kuhn55} in polynomial time; 
	%[TODO]

	%however, to the knowledge of the author, no algorithm exists to more generally compute the $i$'th matching in increasing order of $C^*_m$ (assuming some fixed tie-breaking). This means that matchings have to be exhaustively enumerated and stored in memory to sort them, but this requires $\mathcal{O}(M)$ memory which was found to be more problematic in practice than the time complexity of the sorting, $\mathcal{O}(M\log(M))$.

@@

	This is why in implementation, ordered enumeration is approximated using a priority queue that maintains a subset of matchings of a fixed size $M_{pq}$\footnote{$M_{pq} = 10000$ was used in implementation but this could be much increased at the cost of memory}. Algorithm \ref{pq-algo} outlines the procedure.

@@

	CBM solving all problems with one team (Figure \ref{fig:r-25-1-p} and \ref{fig:r-75-1-p}), problems equivalent to anonymous MAPF instances, is at first surprising but follows directly from the definition of the CBM method. CBM solves single teams as anonymous MAPF instances in polynomial time, meaning that solving a single team reduces to a polynomial-time procedure, namely SSP. Following the results reported in \cite{ma2016}, it is expected that as the number of teams approaches $k$, CBM performance will deteriorate and thus become more similar to the results for exhaustive methods here reported.
	%	Still, some implementations might be more optimized than others so this might be a factor in performance characteristics and in particular in relative performance differences.
	\section{Conclusions and Future work}
	\label{conclusions}
	\label{future}
	The goal of this research was to seek methods for solving MAPFM based on ICTS. In this, the research has succeeded, but the scalability of the algorithms leaves something to be desired. Exhaustive methods, like those discussed in detail in Subsection \ref{exhaustive}, cannot be expected to scale well as teams grow larger due to their complexity, no matter what optimizations are put in place. Nonetheless, as was shown in Section \ref{experiments}, the more advanced exhaustive ICTS algorithms consistently outperform the ICTS-m based algorithms. Compared to the SIC-adaptation of the CBM method \cite{baauw2021}, it was seen that exhaustive methods generally perform relatively poorly if there are fewer teams and fewer obstacles, while being competitive once teams grow smaller and maps become more constrained. Delineating under which conditions exhaustive methods outperform CBM and vice versa is something for further research.

	%\section{Future work}
	A key idea for future research to look into is how to exploit the lower-bound on solution SIC that can be found using the Hungarian Algorithm \cite{kuhn55}. This lower-bound could be used in a branch-and-bound \cite{land1960} search of matchings. The most straightforward way would be to use an algorithm for MAPF at leaf nodes but algorithms for MAPFM could be used to solve (small) branches. Another possibility would be to use this bound to improve ICTS-m by only executing the $k$-agent low-level search for nodes that satisfy this bound. Lower-bounding in ICTS-m could then be further improved by finding the optimal SIC for the anonymous MAPF instance corresponding to each team using a similar approach as \cite{baauw2021}. As long as there is a team that does not meet its bound, branching only on agents within this team is possible without losing optimality.

	A related observation is that the algorithms introduced in this work do not rely on the property of agents and goals having one color. Both exhaustive ICTS and ICTS-m could easily be adapted to support one or more colors (teams) for each agent and goal while still guaranteeing optimality, in contrast to CBM \cite{ma2016} and its SIC adaptation \cite{baauw2021} which specifically use this property to plan single teams in polynomial time. Future research could study the multi-color MAPFM generalization and the application of methods described in this work to that problem.
